---
title: "Creating Start-Stop Data in R"
output: rmarkdown::html_vignette
author: "Robin Denz"
vignette: >
  %\VignetteIndexEntry{Creating Start-Stop Data in R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.width=6.7, fig.height=4.6)
```

# Introduction

A very common task in survival analysis and other types of analyses that involve time-varying covariates is the creation of start-stop datasets. These datasets should include multiple intervals per subject or case, along with one or multiple covariate values that correspond to the value observed during the defined time-intervals. The `MatchTD` package offers a wide range of functions to create and transform such datasets from other data sources (mostly because they are also required as input in its main function, `match_td()`).

The goal of these functions is to make it easy to create valid start-stop datasets and to further process existing ones, even when there are millions of cases and/or a large amount of intervals. The current standard tool to create start-stop data is the `tmerge()` function from the `survival` package, which was not originally meant to handle large datasets. In contrast, the implementations offered here rely exclusively on code written using the `data.table` backend, making them faster and more RAM efficient. Additionally, the functions included in `MatchTD` work well with incomplete information, which often occurs in real data.

# What is Start-Stop Data?

First, let's be a little more clear what exactly constitutes a start-stop dataset. Consider the following example:

```{r}
library(data.table)
library(MatchTD)

data <- data.table(id=c(1, 1, 1, 2, 2, 2, 2),
                   start=c(0, 10, 50, 0, 28, 120, 125),
                   stop=c(10, 50, 112, 28, 120, 125, 213),
                   sex=c("m", "m", "m", "f", "f", "f", "f"),
                   bmi=c(32, 34, 38, 27, 28, 35, 26))
print(data)
```

The dataset shown above consists of two individuals (two distinct `id` values) with multiple recorded time-intervals each (defined by the `start` and `stop` columns). Additionally, it includes information on the `sex` and the `bmi` (Body-Mass-Index) of the individual. In this concrete example, individual 1 is considered male (`sex="m"`) and individual 2 is considered female (`sex="f"`) for the entire observed duration. In other words, `sex` is a **time-fixed** of **time-constant** variable. On the other hand, each of the two individuals have multiple different values in the `bmi` column, making it a **time-varying** or **time-dependent** variable.

More specifically, we would say that the value of `bmi` for individual 1 was 32 from 0 up to and including t = 10. An infinitesimally amount of time later it changed to 34 and stayed on that value until t = 50. This type of coding is known as **right-closed** or **left-open** intervals (because the `stop` value is considered to be in the interval, while the `start` value is not). This is usually denoted as `(start, stop]` and is the usual way these intervals are coded for time-to-event purposes. Throughout this package we will rely exclusively on this type of intervals.

Note that in reality the values of time-dependent variables does not necessarily change instantaneously. To allow a data representation it is however necessary to define some intervals in which the value stays constant. The width of these intervals is usually dependent on the type of data one has access to.

# Overview of Included Functions

The following functions of `MatchTD` may be used to deal with start-stop data:

* `merge_td()`: Merges two or more start-stop datasets
* `subset_start_stop()`: Takes a subset of intervals from a start-stop dataset
* `simplify_start_stop()`: Combines rows from a start-stop dataset when possible
* `add_missing_intervals()`: Adds missing time-intervals to "incomplete" start-stop datasets
* `start_stop2long()`: Creates a long-format dataset from a start-stop dataset
* `long2start_stop()`: Creates a start-stop dataset from a long-format dataset
* `times_from_start_stop()`: Extracts event times from a start-stop dataset

Most users will only need the `merge_td()` function to create a single start-stop dataset from different datasets containing time-dependent information. The other, more advanced, functions are most useful when dealing with messy real data.

# The `merge_td()` function

The most important function to create start-stop data included in this package is the `merge_td()` function. Its' main purpose is to merge two or more datasets that contain information about time-intervals. 

# Generating Start-Stop Datasets from Scratch

## The Input

In many cases the information that should be included in the desired start-stop dataset is stored in different tables that are not already in the start-stop format themselves. For example, suppose we are interested in creating a start-stop dataset including information about the `bmi`, `sex`, `birthyear` and `chemo`-therapy status of different individuals over time. Suppose further that we only have the following three datasets:

```{r}
d_demographics <- data.table(id=c(1, 2),
                             sex=c("m", "f"),
                             birthyear=c(1954, 1962))

d_bmi <- data.table(id=c(1, 1, 1, 2, 2),
                    start=c(25, 190, 256, 78, 235),
                    bmi=c(41, 37, 32, 39, 31))

d_chemo <- data.table(id=c(1, 2),
                      start=c(112, 82))
```

In this example all three datasets contain information about two fictional individuals, `id = 1` and `id = 2`. The dataset `d_demographics` contains some time-constant information about each individual (`sex` and `birthyear`). The dataset `d_bmi` on the other hand, consists of `bmi` measurements of each person taken at a specific point in time. Similarly, the `d_chemo` dataset includes the time at which a person received a chemotherapy. To generate a start-stop dataset from these three tables, we will have to make some decisions on how the resulting dataset should look like. Based on these decisions, we need to augment them a little bit before we can use the functions included in this package.

The general course of action should be:

* 1.) Define singular start-stop datasets for each time-varying variable
* 2.) Merge them together using `merge_td()`

These points are explained in detail below.

## Defining Time-Intervals for each Time-Varying Variable

### BMI

The first question we need to ask is: How do we interpolate between `bmi` measurements? Since we only took measurements at specific times, we don't actually know the value of `bmi` between those measurements. In this case we will simply assume that the value stayed constant until the next measurement (which is what is usually done in practice). Given this decision, we have to create time-intervals that define where the respective `bmi` value applies. I conveniently already name the time variable `start` here to make this a little more obvious. This can be done in three steps.

*1.) Sort the dataset by `id` and `time`*

Using the `data.table` package, this can be done like this (although it is already sorted in this example):

```{r}
setkey(d_bmi, id, start)
```

*2.) Shift the time variable*

Next we again use the `data.table` package to shift the time variable up by one row per `id`, creating the `stop` column:

```{r}
d_bmi[, stop := shift(start, type="lead"), by=id]
```

*3.) Deal with the last row*

Through step 2 we were able to create a `stop` value for each row per person, except the last one. The question in this case is, how long can the last measured value of `bmi` be expected to be valid? In many cases there is some administrative end of the follow-up which can serve to define this point in time. In other cases some assumptions may be reasonable. If neither of these options work, it might be best to just remove those rows. In our example we will consider a final follow-up time of t = 500:

```{r}
d_bmi[is.na(stop), stop := 500]
```

The result looks like this:

```{r}
print(d_bmi)
```

Essentially, this is now a start-stop dataset by itself, which is the exactly the goal of this exercise. Once we have start-stop datasets for each time-varying variable, we can simply merge them using the `merge_td()` function.

Note that it may also be a valid choice to "flip" the intervals. In this case we could have coded the intervals to start at 0 (or some other t) and stay at the first observed value of `bmi` until t = 25 for id = 1. Which option to choose might have serious consequences, but depends entirely on what kind of data you are dealing with.

### Chemo

The next question we need to answer is: What kind of `chemo` variable do we need? There is a unlimited variety of options here. We could, for example, be interested in a sort of variable which includes the information "has received a `chemo` in the last x time units". For exemplary purposes we will first produce a variable that simply states whether someone received a `chemo` in the last 100 time units. This can be done by again defining a fitting `stop` column:

```{r}
d_chemo[, stop := start + 100]
```

We have now created a start-stop dataset again. Secondly, we have to define the value that this variable should have if that condition is met. We could use something like `"yes"` or `1`, but the simplest (and most memory efficient) approach is to use `TRUE`. Since the mini start-stop dataset only includes rows where the condition is met, we can simple set the entire column to this value:

```{r}
d_chemo[, chemo := TRUE]
```

The resulting dataset looks like this:

```{r}
print(d_chemo)
```

Note that there is no need to change the `d_demographics` dataset, because it does not contain any time-varying information.

## Merging the Time-Intervals for each Time-Varying Variable

All that is left to do is call the `merge_td()` function to, well, merge the time-intervals of `bmi` and `chemo`. The following function call may be used:

```{r}
out <- merge_td(d_chemo, d_bmi, by="id", defaults=list(chemo=FALSE),
                constant_vars=d_demographics)
print(out)
```

Again, we simply supply the variable-specific start-stop datasets first and set the `by` argument to the person identifier. Additionally, we set the `defaults` argument to `list(chemo=FALSE)`, which essentially tells the `merge_td()` function that the value of the `chemo` column should be `FALSE` for all time-intervals outside of the ones defined by the `d_chemo` dataset. Additionally, since the information in `d_demographics` is time-independent, it should be supplied to the `constant_vars` argument.



